---
title: 'HW2-EDA with R - Part1: Plotting'
author: "misken"
date: "September 29, 2019"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment

In this assignment you'll do some exploratory data analysis
with R on a dataset of US census information based on the [American Community Survey](https://www.census.gov/programs-surveys/acs/). The version of
this data that we'll be using is from a Kaggle dataset. See
https://www.kaggle.com/muonneutrino/us-census-demographic-data/data and look
at the column metadata for the county level table to see what the various
columns mean. The official documentation for the ACS data can be found at
https://www2.census.gov/programs-surveys/acs/tech_docs/subject_definitions/2017_ACSSubjectDefinitions.pdf.

As we did in class, you'll be creating an R Markdown document to
both prepare a dataset and do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document

Save this file as a new R Markdown document and name it **HW2_EDAwR_[_your last name_].Rmd**. Mine would
be called **HW2_EDAwR_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

Create a new R Project based on this working directory.

**Hacker Extra** - Put this project folder under version control using git. Only
track the .R file, the .Rmd file and your .gitignore file. Create a remote repo
for this homework assignment in your GitHub account. Periodically push your commits
to GitHub. When done with the assignment, paste in the URL of your GitHub project
so that I can go look at it.
  
>Github URL: https://github.com/Alice4ever/HW2

## Step 3 - Explore data folder

Inside of the main project folder you'll find the data files for this assignment in a subfolder called `data/`.

- **acs2017_census_tract_data.csv**
- **acs2017_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset. The two files,
**county.csv** and **county_MI.csv**, are files that I created. The first
contains county level data from all states and the other just data from
Michigan. The R script, **census_data_prep.R** was used to create these files
and to merge information from the **us_census_bureau_regions_and_divisions.csv**
file with the county data. **You do NOT need to rerun the census_data_prep.R file.** I'm
including it and encouraging you to take a look through it to get a sense of what a
typical data prep script might look like. We'll be learning to do all of the things
shown in that file.

You might want to use shell commands or a text editor like Geany to snoop the files a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class, you should
use a combination of markdown text (be concise, no need to write tons of text) 
to explain what you are doing and R code chunks to actually do it.
When you are done, use the "Knit" button to generate an HTML file from your R Markdown.

Note that if you have errors in your R code, **knitr** will stop at the error
and won't create an html file. Sooo, you should comment out R code that has
errors and write a little comment stating that it's an error. Keep doing this
until you successfully knit an html file based on your R Markdown document.
You'll be submitting both the html file as well as the Rmd file.

You'll be submitting BOTH the completed R Markdown file as well as the generated HTML file. You'll notice that I've left some empty code chunks
as well as some partially filled in code chunks. Of course, feel free to
add more code chunks as needed.

Let's load a few libraries we'll need:

```{r}
library(ggplot2)
```


### Problem 1 - Get the data into a dataframe

You should read the **county.csv** file into a data frame named county_df and
**county_mi_df.csv** into a data frame named county_mi_df.

```{r read_countydata}
county_df <- read.table("data/county.csv",
                      header=TRUE, sep=",",
                      stringsAsFactors=TRUE)
county_mi_df <- read.table("data/county_MI.csv",
                      header=TRUE, sep=",",
                      stringsAsFactors=TRUE)

```

* What does each row represent?

> Each row represents an observation/record.

Check the structure of each data.frame and use the `summary` function to begin
exploring the data. 

* Are there any NA values? 
* What differences do you notice between the structure of the two data frames.


```{r}
str(county_df)
str(county_mi_df)
summary(county_df)
summary(county_mi_df)
```

> ChildPoverty in data frame county_df has 1 NA value.
county_mi_df has one more column named "Peninsula" than county_df. It has less obersavations because most of the data is a subset data from county_df and it is only for Michigan state.

**Unless otherwise specified, you should use the ggplot2 package for all of the
following plots.**

### Problem 2 - Basic plots for Income

Create a histogram and a kernel density plot for Income and for IncomePerCap for the entire US (i.e. use the `county_df` dataframe.) The Income field is median income for the county and IncomePerCap is aggregate income divided by population size, for the county.

In addition, compute and display the following statistics for Income and IncomePerCap:

* mean and median
* standard deviation
* min, max and range
* 95th and 99th percentile
* IQR
* coefficient of variation


```{r summary_stats}
# summary stats for Income
mean_income <- mean(county_df$Income)
mean_income

median(county_df$Income)

sd_income <- sd(county_df$Income)
sd_income

min_income <- min(county_df$Income)
min_income

max_income <- max(county_df$Income)
max_income

range_income <- max_income - min_income
range_income

quantile(county_df$Income,0.95)
quantile(county_df$Income,0.99)

IQR_income <- quantile(county_df$Income,0.75) - quantile(county_df$Income,0.25)
IQR_income

cv_income <- sd_income / mean_income
cv_income

# summary stats for IncomePerCap
mean_IncomePerCap<- mean(county_df$IncomePerCap)
mean_IncomePerCap

median(county_df$IncomePerCap)

sd_IncomePerCap<- sd(county_df$IncomePerCap)
sd_IncomePerCap

min_IncomePerCap<- min(county_df$IncomePerCap)
min_IncomePerCap

max_IncomePerCap<- max(county_df$IncomePerCap)
max_IncomePerCap

range_IncomePerCap<- max_IncomePerCap - min_IncomePerCap
range_IncomePerCap

quantile(county_df$IncomePerCap,0.95)
quantile(county_df$IncomePerCap,0.99)

IQR_IncomePerCap <- quantile(county_df$IncomePerCap,0.75) - quantile(county_df$IncomePerCap,0.25)
IQR_IncomePerCap

cv_IncomePerCap <- sd_IncomePerCap / mean_IncomePerCap
cv_IncomePerCap


```

```{r histo_dens}
options(scipen = 999)
ggplot(data = county_df) +
  geom_histogram(mapping = aes(x=Income))
ggplot(data = county_df) +
  geom_histogram(mapping = aes(x=IncomePerCap))
ggplot(data= county_df) + 
  geom_density(aes(x=Income))
ggplot(data= county_df) + 
  geom_density(aes(x=IncomePerCap))
```


Write a short paragraph describing the distribution of of these two variables.

> Your interpretation of the summary stats and plots above...

> These two distributions are skewed to right slightly with a single peak. The right tail (larger values) is much longer than the left tail (small values). This indicates the bulk of the Income/IncomePerCap are small/medium, with a few that are much larger than the rest. There are even outliers at the right end which fall outside the overall histogram pattern indicating the presence of a very large value.

Repeat the histogram for Income but make the following style changes to the plot.

* Add a title such as "Distribution of median income for US counties"
* Change y-axis label to "Number of counties"
* Change x-axis label to "Median income"
* Change the colors of the bars to something other than the default

```{r histo_stylechanges}
ggplot(data = county_df) + 
  geom_histogram(mapping = aes(x=Income), fill="blue", color="black") +
  labs(title="Distribution of median income for US counties", x="Median income", y="Number of counties")
```

Now create a set of histograms for Income in which you facet by Division.

```{r histos_division}
ggplot(county_df) +
  geom_histogram(aes(x=Income)) +
  facet_wrap(~Division)
```

Since the number of counties varies by Division, create a new version of the plot above but modify so that the y-axis shows frequency percentages instead of
counts. Hint: `..density..`

```{r histos_division_density}
ggplot(county_df) +
 geom_histogram(aes(x=Income, y=..density..)) +
 facet_wrap(~Division)
```


Now create boxplots and violin plots for the same two variables. However, you'll
see that for both of these plot types, ggplot wants an `x` aesthetic to act as a grouping variable. Use the `region` field as your `x` aesthetic. Note anything of interest revealed by these plots.

```{r box_violin}
# Boxplots and violin plots
options(scipen = 999)
  ggplot(county_df) +
  geom_boxplot(aes(x=Region, y=Income), 
               fill="#FF9999", colour="black")
  ggplot(county_df) +
  geom_violin(aes(x=Region, y=Income), 
              fill="steelblue", colour="black")
  
  ggplot(county_df) +
  geom_boxplot(aes(x=Region, y=IncomePerCap), 
               fill="#FF9999", colour="black")
  ggplot(county_df) +
  geom_violin(aes(x=Region, y=IncomePerCap), 
              fill="steelblue", colour="black")
```

> Your interpretation ...

>Income and IncomePerCap shows similar pattern of boxplots and violin plots with respect to different regions. 
The outliers in boxplots are corresponded to the thinnest sides of violin plots.
We can only read median, range and outliers from boxplots, but violin plots tell more data information. The widest part shows the high frequency data and the thinest side shows least frequency data. For example, from the income box plot, we can see Northeast has a higer median than Midwest, but from the violin plot, at the same $50000 level, we can see Midwest has a relatively higher frequency.


Repeat the histogram of Income but only for counties with less than 500,000 people. Do NOT use
dplyr to do the filtering - instead, use boolean indexing (see intro to dataframes Rmd file).

```{r histo_lt1.5m1}
is.county.ok <- county_df$TotalPop < 500000
 ggplot(county_df[!is.county.ok,]) +
  geom_histogram(aes(x=Income), fill="blue", color="black") +
  labs(title="Distribution of median income for county population less than 500000", x="Median income", y="Number of counties")


```

**Hacker Extra** - Also create a kernel density plot of Income overlaid on a histogram of
Income. 

```{r hackerxtra_kdoverlay}
ggplot(county_df[!is.county.ok,]) +
  geom_histogram(aes(x=Income,y=..density..), fill="blue", color="black") +
  labs(title="Distribution of median income for county population less than 500000", x="Median income")+
  geom_density((aes(x=Income)),alpha=.2, fill="#FF6666")

```

### Problem 3 - How do Michigan's peninsulas compare?

Create violin plots showing Income by peninsula. Make sure
your graph has a title and good axis labels.

```{r vp_inc_by_peninsula}
ggplot(county_mi_df) + 
  geom_violin(aes(x= peninsula, y= Income),fill="steelblue", colour="black") +
   labs(xlab="Peninsula", title="Michian Income By Peninsula")


```

Do the peninsulas 
differ with respect to distribution of median income over the counties?

> Your discussion...

> Yes, they do differ with repect to distribution of median income.

> Compared with Lower peninsula,the distribution of median income in Upper peninsula looks more like a normal distribution because the both sides of the vilion plot are having a similar shape. This indicated that median income over the counties of upper peninsula are having less diversities. They are more close to each other, all clustering within 35000-50000.

> On the other hand, the violin plot of LP has a longer, thinner shape. The thinner upper side indicates that there are some outliers, a small group of large numbers are pulling up the median income. The median income varies more widely from below 35000 up to 80000. All the higher median income (>50000) are in lower peninsulas.

Create several boxplots and/or violin plots showing how the peninsulas compare with respect to
some of the other variables. Summarize your findings in a few short paragraphs.

```{r boxplots_upvlp}
ggplot(county_mi_df) + 
  geom_violin(aes(x= peninsula, y= TotalPop),fill="steelblue", colour="black") +
   labs(xlab="Peninsula", title="Michian Population By Peninsula")+
    scale_y_log10()
ggplot(county_mi_df) + 
  geom_boxplot(aes(x= peninsula, y= TotalPop),fill="steelblue", colour="black") +
   labs(xlab="Peninsula", title="Michian Population By Peninsula")+
  scale_y_log10()
ggplot(county_mi_df) + 
  geom_violin(aes(x= peninsula, y= MeanCommute),fill="steelblue", colour="black") +
   labs(xlab="Peninsula", title="Michian Mean Commute Time By Peninsula")
ggplot(county_mi_df) + 
  geom_violin(aes(x= peninsula, y= Construction),fill="steelblue", colour="black") +
   labs(xlab="Peninsula", title="Michian Construction By Peninsula")
ggplot(county_mi_df) + 
  geom_violin(aes(x= peninsula, y= Poverty),fill="steelblue", colour="black") +
   labs(xlab="Peninsula", title="Michian Poverty Distribution By Peninsula")
ggplot(county_mi_df) + 
  geom_violin(aes(x= peninsula, y= Professional),fill="steelblue", colour="black") +
   labs(xlab="Peninsula", title="Michian Professional Distribution By Peninsula")
```

> Your discussion ...

> Population: The majority of Upper Peninsula counties have smaller counties with total population around 10000 and the biggest county's population is less than 100000. While in lower peninsula, smallest county is almost 10000 and the majority counties are around 50000. A few bigger county are between 10000 and up to 1000000. We can say compared with lower peninsula, upper peninsula has relatively smaller counties and county size is no big difference.

> Mean Conmmute Time: Although the majority of commute time is similar to each other.
lower peninsula counties have a relatively longer commute time than upper peninsula. Their shortest and longest commutetime are both longer than the ones in upper peninsula.

> With respect to contruction, poverty and professional, lower peninsula has a wider range of different levels over the counties than upper peninsula.The majotiy of upper peninsula has higher poverty level, less professional level and more construction.

### Problem 4 - Basic bar chart

Create a bar chart showing the number of counties by State in the US. Make sure
your graph has a title and good axis labels. Figure out how to format the
x-axis labels so that they don't obscure each other.

```{r bar_counties_by_state}
ggplot(county_df) +
  geom_bar(aes(x=State))+ 
  ggtitle("Counties By State") +
  theme(axis.text.x= element_text(angle = 90, hjust = 1))

```

**Hacker Extra** Modify the above bar plot to be horizontal and have bars sorted from longest to shortest. In addition, map the bar color to the Region variable.

```{r bar_counties_by_state_hackerxtra, fig.height=6}

library(dplyr)
county_df %>% 
  count(State, Region) %>% 
  ggplot(aes(x = reorder(State, n), y = n, fill=Region))+
  geom_col()+
  coord_flip()+
  ggtitle("Counties By State")+
  ylab("Count of Counties")
```

### Problem 5 - Basic scatter plots

Create a scatter plot using the Michigan only data frame comparing
IncomePerCap and MeanCommute.

```{r xy_inc_commute}
ggplot(data = county_mi_df) + 
  geom_point(aes(x=MeanCommute, y=IncomePerCap))
```

Create a modified version of the above scatter plot in which the point color
is mapped to peninsula and the size is mapped to total population. In addition,
create a similar plot comparing Income and Professional.

```{r xy_inc_commute_peninsula}
ggplot(data = county_mi_df) + 
  geom_point(aes(x=MeanCommute, y=IncomePerCap, color=peninsula, size=TotalPop))

ggplot(data = county_mi_df) + 
  geom_point(aes(x=Professional, y=Income, color=peninsula, size=TotalPop))
```

> Your thoughts on the relationship between these variables?

> By examing the scatter plot of IncomePerCap and MeanCommute, we can see that the plotting does not show a very obvious uphill/downhill pattem from left to right. So the MeanCommute does not relate to IncomePerCap very much, longer MeanCommute time does not indicate a better IncomePerCap.

> By examing the plot size and color,we can see LP Peninsula has more TotalPop and people there has longer MeanCommute time and more IncomePerCap; On the other hand, Upper Peninsula's average commute time is less than 25min and IncomePerCap is less than $30000. Their population is smaller too.
> The scatter plot of Income and Professional shows a strong positive relationship between these two variables, which indicates people having higher professional level has higher income. Still, Upper Peninsula with a smaller population has a lower income(no higher value >50000) and all the professional levlel is below 40. Several higher profrofessional level sample are all belong to Lower Peninsula.


Now, create a few more scatter plots using variables of your choosing in which
you facet by peninsula as well. How do the relationships compare across the
peninsulas?


```{r xy_faceted_peninsula1}
ggplot(data = county_mi_df) + 
  geom_point(aes(x=MeanCommute, y=IncomePerCap, colour= peninsula))+
  facet_wrap(~peninsula)
```


> For Lower peninsula, scatterplots of MeanCommute and IncomePerCap still does not indicate a positive relationship between them, while for upper peninsula, it looks like with the increase of mean commute time, the income per cap increases(except some outliers.)

```{r xy_faceted_peninsula2}
ggplot(data = county_mi_df) + 
  geom_point(aes(x=Construction, y=MeanCommute,colour= peninsula))+
facet_wrap(~peninsula)
```

> this plots showed a slight positive relationship between construction and meancommute in both peninsulas. It looks like upper peninsula has a relatively stronger correlation between these two variables.

```{r xy_faceted_peninsula3}
ggplot(data = county_mi_df) + 
  geom_point(aes(x=Professional, y=Income,colour= peninsula))+
facet_wrap(~peninsula)
```


> This scatter plot shows in both peninsulas, a better professional level indicates a better income. Upper peninsular has a wider range of Income as well as Professioal level than lower peninsular.

```{r xy_faceted_peninsula4}
ggplot(data = county_mi_df) + 
  geom_point(aes(x=Professional, y=ChildPoverty,colour= peninsula))+
facet_wrap(~peninsula)

```


> This shows in  both peninsulas, a better professional level indicates less child poverty. Likewise, Upper peninsular has more varieties of child poverty  as well as Professioal level.



Now use the full US data frame, county_df.

Create a scatter plot of TotalPop by IncomePerCap.

```{r xy_pop_inc}
options(scipen = 999)
ggplot(data = county_df) + 
  geom_point(aes(x=IncomePerCap, y=TotalPop))
```


The widely varying values make this difficult to interpret in terms of looking
for patterns. So, figure out a way to either change the axis scales or 
the plotted data itself to use a base 10 logarithmic scale (or transform).
Write a short paragraph on why one might want to use a log scale or log
transform of data when plotting.


```{r log10}
ggplot(county_df) +
  geom_point(aes(x=IncomePerCap, y=TotalPop)) + 
   ylab("log(TotalPop)") + 
    scale_y_log10()
  
```

> Discussion ...

> In the first plot, the majority of points were clustered on the bottom part of the graph because Y axis scale is wide, which makes it hard to visulatize the plotting pattern and relationship between two variables. To improve this, one might use log transformation to either transform axis scale or transform the column data.

> In my plot,I transformed Y axis scale to make the plotting pattern more clearer. It seems like these two variables have a positive relationship.

## Deliverables

Make sure you knit your Rmd file. As discussed earlier, if you have errors, your document won't knit. Comment out error lines and put a comment in saying that you couldn't get the line to work. After your file is knitted into an html file, simply create a compressed archive (either tar.gz or .zip) of your project folder and upload the compressed archive file into Moodle.